Decoding the Amplitude-Modulated Part the DCF77 Longwave Time Signal by using the Goertzel Algorithm

### 2024-08-18 paper/main.tex
Progress in explaining the structure of the receiver

@@

and therefore a focus is put on explaining this part of the DCF77 signal.

@@

\section{Principle of the receiver}
To understand the principle of the receiver, it is compared to the common structure that is used in radio controlled clocks.
\begin{figure}[htbp]
    \centering
    \resizebox{0.48\textwidth}{!}{
        \begin{tikzpicture}
            % Block styles
            \tikzstyle{block} = [rectangle, draw, text width=4.5em, text centered, minimum width=6em, minimum height=4em]
            \tikzstyle{block_dashed} = [rectangle, draw, text width=2em, text centered, minimum width=4em, minimum height=4em, dashed]
            % Input and output
            \begin{scope}[shift={(-3,0)}] % Adjust the position of the antenna
                \coordinate (antenna) at (0,0);
                \draw[thick] (0,0) -- (0,1); % Antenna pole
                \draw[thick] (-0.25,1) -- (0,0.75); % Antenna horizontal line
                \draw[thick] (0.25,1) -- (0,0.75); % Antenna horizontal line
            \end{scope}
            \node[block, right=of antenna, minimum width=4em, minimum height=4em] (bpf) {BPF};
            \node[block, right=of bpf] (rfamp) {RF Amp};
            \node[block, right=of rfamp] (adc) {ADC};
            \node[block, below=of adc] (timer) {Timer};
            \node[block, right=of adc] (goertzel) {Goertzel Algorithm};
            \node[block, right=of goertzel] (det) {Detector};
            % Connections
            \draw[thick] (antenna) -- (bpf);
            \draw[->] (bpf) -- (rfamp); % Connection from antenna to RF amplifier
            \draw[->] (rfamp) -- (adc);
            \draw[->] (timer) -- (adc);
            \draw[->] (adc) -- (goertzel);
            \draw[->] (goertzel) -- (det);

            \node[draw, dashed, fit={(antenna) (bpf)}, inner sep=12pt] (tank_circuit) {};
            \node[below=0.5em of tank_circuit, align=center] {Tuned Tank Circuit};
        \end{tikzpicture}
    }
\end{figure}
\FloatBarrier\noindent
Test

@@

            %f_s and their multiples
            \foreach \x in {-3,-2,-1,1,2,3}{
                \draw[-> , thick] ($\x*(fs)$) -- ++(0,1.5);
                \draw[very thick] ($\x*(fs) + (0,-0.2)$) -- ++(0,0.4);
                \node at ($\x*(fs) + (0,-0.4)$) {$\x f_{s}$};

                \draw[thick]
                ($(fc) + (-0.12,0) + \x*(fs)$) -- % Left line start
                ++(0,1) -- %Left line up
                ++(0.24,0.5) -- %Roof
                ++(0,-1.5) -- %Right line down
                cycle; % Close the shape
                \draw[very thick] ($(fc) + (0, -0.2) + \x*(fs)$) -- ++(0, 0.4); %Line x-axis
                \node at ($(fc) + (0, -0.8) + \x*(fs)$) {$f_{c} \pgfmathprintnumber[print sign]{\x} f_{s}$}; %Index x-axis

                %DCF77 at -f_c
                \draw[thick]
                ($-1*(fc) + (-0.12,0) + \x*(fs)$) -- % Left line start
                ++(0,1.5) -- %Left line up
                ++(0.24,-0.5) -- %Roof
                ++(0,-1) -- %Right line down
                cycle; % Close the shape
                \draw[very thick] ($-1*(fc) + (0, -0.2) + \x*(fs)$) -- ++(0, 0.4); %Line x-axis
                \node at ($-1*(fc) + (0, -0.8) + \x*(fs)$) {$-f_{c} \pgfmathprintnumber[print sign]{\x} f_{s}$}; %Index x-axis
            };
        \end{tikzpicture}
    } 
\end{figure*}

@@

\begin{thebibliography}{00}
\bibitem{b1} -, ``DCF77,'' 2024, Wikipedia entry. [Online]. Available: https://de.wikipedia.org/wiki/DCF77
\bibitem{b2} D. Priester and P. Hetzel and A. Bauch, ``Zeit- und Normalfrequenzverbreitung mit DCF77,'' PTB Mitteilungen, 2004, [Online]. Available: https://www.ptb.de/cms/fileadmin/internet/fachabteilungen/abteilung\_4/4.4\_zeit\_und\_frequenz/4.42/dcf77.pdf
\bibitem{b3} D. Priester and P. Hetzel and A. Bauch, ``Zeit- und Normalfrequenzverbreitung mit DCF77: 1959-2009 und darüber hinaus,'' PTB Mitteilungen, 2009, [Online]. Available: https://www.ptb.de/cms/fileadmin/internet/fachabteilungen/abteilung\_4/4.4\_zeit\_und\_frequenz/pdf/2009\_Bauch\_PTBM\_\_DCF77.pdf
\bibitem{b4} P. Hetzel, ``Time dissemination via the LF transmitter DCF77 using a pseudo-random phase-shift keying of the carrier,'' PTB Braunschweig, 1988, [Online]. Available: https://www.ptb.de/cms/fileadmin/internet/fachabteilungen/abteilung\_4/4.4\_zeit\_und\_frequenz/pdf/5\_1988\_Hetzel\_-\_Proc\_EFTF\_88.pdf
\bibitem{b5} -, ``DCF77 webpage,'' 2024, Official page of the PTB [Online]. Available: https://www.ptb.de/cms/en/ptb/fachabteilungen/abt4/fb-44/ag-442/dissemination-of-legal-time/dcf77.html
%\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
%\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
%\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
%\bibitem{b6} Y. Yorozu, M. Hirano, K. Oka, and Y. Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl. J. Magn. Japan, vol. 2, pp. 740--741, August 1987 [Digests 9th Annual Conf. Magnetics Japan, p. 301, 1982].
%\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
%\bibitem{b9} S. Liu, ``Wi-Fi Energy Detection Testbed (12MTC),'' 2023, gitHub repository. [Online]. Available: https://github.com/liustone99/Wi-Fi-Energy-Detection-Testbed-12MTC
%\bibitem{b10} ``Treatment episode data set: discharges (TEDS-D): concatenated, 2006 to 2009.'' U.S. Department of Health and Human Services, Substance Abuse and Mental Health Services Administration, Office of Applied Studies, August, 2013, DOI:10.3886/ICPSR30122.v2
%\bibitem{b11} K. Eves and J. Valasek, ``Adaptive control for singularly perturbed systems examples,'' Code Ocean, Aug. 2023. [Online]. Available: https://codeocean.com/capsule/4989235/tree
\end{thebibliography}

>>>>

\usepackage{ifthen}

@@

and therefore a focus is put on explaining this part of the DCF77 signal. The following explainations in this paper therefore refer explicitly to the amplitude-modulated part.

@@

\section{Principle of the receiver}
As the seen in the description of the DCF77 signal, retrieving the the binary time signal from the amplitude-modulated part is more or less trivial.
Conventional applications execute this task by using analog receiving circuits, usually by using a simple amplitude modulation receiver together with a
tuned ferrite-rod antenna, acting as a Bandpass Filter (BPF), to filter out the irrelevant part of the spectrum. 
Moving the receiver into the digital domain allows for more advanced techniques to retrieve this binary time signal, rather than demodulating the signal with a
diode-detector inside the amplitude modulation receiver.
Because the information is contained only in the amplitude of the signal, and therefore inside the signal strength at the receiver, it can also be obtained by doing a
continuous analysis of the spectrum at the carrier's frequency of \SI{77.5}{\kilo\hertz}.
\subsection{The Goertzel Algorithm}
The standard approach to analyze a spectrum in the digital domain after sampling is the well-known Discrete Fourier Transform (DFT), implemented by using
the Fast Fourier Transform (FFT). This yields an analysis of the complete spectrum, producing $\frac{N}{2} + 1$ relevant results for equally distributed frequencies
("frequency bins") within range from \SI{0}{\hertz} to $\frac{f_{s}}{2}$ where $N$ is the number of samples and $f_{s}$ the sampling frequency \cite{b6}.
The signal strength, and therefore the binary time signal of the amplitude-modulated part can then be examined by observing the value of the curresponding frequency bin
over time. Because of only one frequency bin being relevant, it is obvious that a full FFT produces a lot of overhead and redundant calculation operations in this use-case. 
This overhead can be avoided by using the Goertzel algorithm instead of the FFT, which is able to calculate the value of one specific frequency bin with a smaller
amount of needed calculations.
It was invented in 1958 by Gerald Goertzel and since then most commonly used in the detection tones employed in Dual-tone Multi-frequency (DTMF) signaling \cite{b7}.
The working principle of the algorithm can be percieved as a second-order Infinite Impulse Response (IIR) filter where the results $y[n-1]$ and $y[n-2]$ of the output
are stored in states $Q_{1}$ and $Q_{2}$, so they are still available after the execution of the filter.
\begin{equation}
    y[n] = x[n] + 2 \cdot cos(\omega_{g}) \cdot y[n-1] - y[n-2]
\end{equation}
From these equations, the following structure can be derived:
\begin{figure}[htbp]
    \centering
    \resizebox{0.4\textwidth}{!}{
        \begin{tikzpicture}[auto, node distance=2cm]
            %Defining a cross set for the multiplication elements
            \tikzset{cross/.style={path picture={ 
            \draw[black]
            (path picture bounding box.south east) -- (path picture bounding box.north west) (path picture bounding box.south west) -- (path picture bounding box.north east);
            }}}

          % Define block styles
            \tikzstyle{delay} = [draw, rectangle, minimum height=2em, minimum width=2em]
            \tikzstyle{sum} = [draw, circle, minimum size=2em, node distance=1.5cm]
            \tikzstyle{mult} = [draw, circle, minimum size=2em, node distance=1.5cm, cross]
            \tikzstyle{state} = [draw, rectangle, rounded corners, minimum height=2em, minimum width=2em]

            % Nodes
            \node at (-2, 0) (x) {$x[n]$};
            \node[sum] at (0,0) (sum1) {+};
            \node[sum] at (0,-2) (sum2) {+};
            \node[mult] at (2,-2) (mult1) {};
            \node[mult] at (2,-4) (mult2) {};
            \node[delay] at (4,-1) (del1) {$z^{-1}$};
            \node[delay] at (4,-3) (del2) {$z^{-1}$};
            \node[state] at (6,-2) (st1) {$Q_{1}$};
            \node[state] at (6,-4) (st2) {$Q_{2}$};
            \node at (6, 0) (y) {$y[n]$};

            \draw[->] (x.east) -- (sum1.west);
            \draw[->] (sum1.east) -- (y.west);
            \draw[->] (sum2.north) -- (sum1.south);
            \draw[->] (mult1.west) -- (sum2.east);
            \draw[->] (mult2.west) -- (0, -4) -- (sum2.south);
            \draw[->] (4, 0) -- (del1.north);
            \draw[->] (del1.south) -- (del2.north);
            \draw[->] (del2.south) -- (4,-4) -- (mult2.east);
            \draw[->] (4,-2) -- (mult1.east);
            \draw[->] (2, -1.25) -- (mult1.north);
            \draw[->] (2, -3.25) -- (mult2.north);
            \draw[->] (4, -2) -- (st1.west);
            \draw[->] (4, -4) -- (st2.west);
            \node at (2, -1) {$2 \cdot cos(\omega_{g})$};
            \node at (2, -3) {$-1$};
            \node[draw, shape=circle, fill=black, scale=0.3] at (4,0) {};
            \node[draw, shape=circle, fill=black, scale=0.3] at (4,-2) {};
            \node[draw, shape=circle, fill=black, scale=0.3] at (4,-4) {};
        \end{tikzpicture}
    }
\end{figure}
\FloatBarrier\noindent

The constant $2 \cdot cos(\omega_{g})$ is computed before the execution. It is calculated based on $\omega_{g}$ and defines the frequency to be analyzed, dependent on the sampling frequency $f_{s}$, the
frequency to analyze $f_{a}$ and the number of input samples $N$.
\begin{equation}
    k = \left\lfloor \frac{N \cdot f_{a}}{f_{s}} \right\rceil
\end{equation}
\begin{equation}
    w_g = \frac{2\pi \cdot k}{N}
\end{equation}

\begin{figure}[htbp]
    \centering
    \resizebox{0.48\textwidth}{!}{
        \begin{tikzpicture}
            % Block styles
            \tikzstyle{block} = [rectangle, draw, text width=4.5em, text centered, minimum width=6em, minimum height=4em]
            \tikzstyle{block_dashed} = [rectangle, draw, text width=2em, text centered, minimum width=4em, minimum height=4em, dashed]
            % Input and output
            \begin{scope}[shift={(-3,0)}] % Adjust the position of the antenna
                \coordinate (antenna) at (0,0);
                \draw[thick] (0,0) -- (0,1); % Antenna pole
                \draw[thick] (-0.25,1) -- (0,0.75); % Antenna horizontal line
                \draw[thick] (0.25,1) -- (0,0.75); % Antenna horizontal line
            \end{scope}
            \node[block, right=of antenna, minimum width=4em, minimum height=4em] (bpf) {BPF};
            \node[block, right=of bpf] (rfamp) {RF Amp};
            \node[block, right=of rfamp] (adc) {ADC};
            \node[block, below=of adc] (timer) {Timer};
            \node[block, right=of adc] (goertzel) {Goertzel Algorithm};
            \node[block, right=of goertzel] (det) {Detector};
            % Connections
            \draw[thick] (antenna) -- (bpf);
            \draw[->] (bpf) -- (rfamp); % Connection from antenna to RF amplifier
            \draw[->] (rfamp) -- (adc);
            \draw[->] (timer) -- (adc);
            \draw[->] (adc) -- (goertzel);
            \draw[->] (goertzel) -- (det);

            \node[draw, dashed, fit={(antenna) (bpf)}, inner sep=12pt] (tank_circuit) {};
            \node[below=0.3em of tank_circuit, align=center] {Tuned ferrite-rod antenna};

            \node[draw, dashed, fit={(adc) (timer) (goertzel) (det)}, inner sep=12pt] (microcontroller) {};
            \node[below=0.3em of microcontroller, align=center] {Microcontroller};
        \end{tikzpicture}
    }
\end{figure}
\FloatBarrier\noindent
Test

@@

            %f_s their result of the convulutions
            \foreach \x in {-3,-2,-1,1,2,3}{
                \draw[-> , thick] ($\x*(fs)$) -- ++(0,1.5);
                \draw[very thick] ($\x*(fs) + (0,-0.2)$) -- ++(0,0.4);
                \node at ($\x*(fs) + (0,-0.4)$) {$\x f_{s}$};

                \draw[thick]
                ($(fc) + (-0.12,0) + \x*(fs)$) -- % Left line start
                ++(0,1) -- %Left line up
                ++(0.24,0.5) -- %Roof
                ++(0,-1.5) -- %Right line down
                cycle; % Close the shape
                \draw[very thick] ($(fc) + (0, -0.2) + \x*(fs)$) -- ++(0, 0.4); %Line x-axis
                \ifthenelse{\x = -3}{
                    \node at ($(fc) + (0.3, -0.8) + \x*(fs)$) {$f_{c} \pgfmathprintnumber[print sign]{\x} f_{s}$}; %Index x-axis
                }{
                    \node at ($(fc) + (0, -0.8) + \x*(fs)$) {$f_{c} \pgfmathprintnumber[print sign]{\x} f_{s}$}; %Index x-axis
                }

                %DCF77 at -f_c
                \draw[thick]
                ($-1*(fc) + (-0.12,0) + \x*(fs)$) -- % Left line start
                ++(0,1.5) -- %Left line up
                ++(0.24,-0.5) -- %Roof
                ++(0,-1) -- %Right line down
                cycle; % Close the shape
                \draw[very thick] ($-1*(fc) + (0, -0.2) + \x*(fs)$) -- ++(0, 0.4); %Line x-axis
                \ifthenelse{\x = 3}{
                    \node at ($-1*(fc) + (-0.3, -0.8) + \x*(fs)$) {$-f_{c} \pgfmathprintnumber[print sign]{\x} f_{s}$}; %Index x-axis
                }{
                    \node at ($-1*(fc) + (0, -0.8) + \x*(fs)$) {$-f_{c} \pgfmathprintnumber[print sign]{\x} f_{s}$}; %Index x-axis
                }
            };
        \end{tikzpicture}
    } 
\end{figure*}

@@

\begin{thebibliography}{00}
\bibitem{b1} -, ``DCF77,'' 2024, Wikipedia entry. [Online]. Available: https://de.wikipedia.org/wiki/DCF77
\bibitem{b2} D. Priester and P. Hetzel and A. Bauch, ``Zeit- und Normalfrequenzverbreitung mit DCF77,'' PTB Mitteilungen, 2004, [Online]. Available: https://www.ptb.de/cms/fileadmin/internet/fachabteilungen/abteilung\_4/4.4\_zeit\_und\_frequenz/4.42/dcf77.pdf
\bibitem{b3} D. Priester and P. Hetzel and A. Bauch, ``Zeit- und Normalfrequenzverbreitung mit DCF77: 1959-2009 und darüber hinaus,'' PTB Mitteilungen, 2009, [Online]. Available: https://www.ptb.de/cms/fileadmin/internet/fachabteilungen/abteilung\_4/4.4\_zeit\_und\_frequenz/pdf/2009\_Bauch\_PTBM\_\_DCF77.pdf
\bibitem{b4} P. Hetzel, ``Time dissemination via the LF transmitter DCF77 using a pseudo-random phase-shift keying of the carrier,'' PTB Braunschweig, 1988, [Online]. Available: https://www.ptb.de/cms/fileadmin/internet/fachabteilungen/abteilung\_4/4.4\_zeit\_und\_frequenz/pdf/5\_1988\_Hetzel\_-\_Proc\_EFTF\_88.pdf
\bibitem{b5} -, ``DCF77 webpage,'' 2024, Official page of the PTB [Online]. Available: https://www.ptb.de/cms/en/ptb/fachabteilungen/abt4/fb-44/ag-442/dissemination-of-legal-time/dcf77.html
\bibitem{b6} R. G. Lyons, ``Understanding Digital Signal Processing,'' Third Edition, Prentice Hall, 2010.
\bibitem{b7} -, ``Goertzel algorithm,'' 2024, Wikipedia entry. [Online]. Available: https://en.wikipedia.org/wiki/Goertzel\_algorithm
%\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
%\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
%\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
%\bibitem{b6} Y. Yorozu, M. Hirano, K. Oka, and Y. Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl. J. Magn. Japan, vol. 2, pp. 740--741, August 1987 [Digests 9th Annual Conf. Magnetics Japan, p. 301, 1982].
%\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
%\bibitem{b9} S. Liu, ``Wi-Fi Energy Detection Testbed (12MTC),'' 2023, gitHub repository. [Online]. Available: https://github.com/liustone99/Wi-Fi-Energy-Detection-Testbed-12MTC
%\bibitem{b10} ``Treatment episode data set: discharges (TEDS-D): concatenated, 2006 to 2009.'' U.S. Department of Health and Human Services, Substance Abuse and Mental Health Services Administration, Office of Applied Studies, August, 2013, DOI:10.3886/ICPSR30122.v2
%\bibitem{b11} K. Eves and J. Valasek, ``Adaptive control for singularly perturbed systems examples,'' Code Ocean, Aug. 2023. [Online]. Available: https://codeocean.com/capsule/4989235/tree
\end{thebibliography}
